Understanding both enhanced and impaired learning with enhanced plasticity: a saturation hypothesis (supplementary material)

### 2016-06-09 MHCpaper/MHCsupp.tex
updated MHC supp pagination
attempted spectra linitialisation
typos in mhc supp
rejigging MHC supp

% -*- TeX -*- -*- UK -*-
% ----------------------------------------------------------------
\documentclass[10pt]{article}
%\usepackage[printfigures]{figcaps}
%\usepackage[nolists,nomarkers]{endfloat}
%Preamble:
\usepackage{a4wide}
\input{sl_preamble.tex}
\hypersetup{colorlinks,linkcolor=blue,citecolor=teal,urlcolor=teal}
\input{sl_graphics_preamble.tex}
\graphicspath{{"Figures/"}}
%
\usepackage{array}
\newcounter{tablerow}[table]
\newcommand{\tablerowautorefname}{row}
\newenvironment{tabularn}[1]{\begin{tabular}{#1>{\refstepcounter{tablerow}}l}}{\end{tabular}}
%\usepackage{etoolbox}
%\AtBeginEnvironment{tabular}{\setcounter{tablerow}{0}}
\renewcommand{\figureautorefname}{Supplementary Figure}
% ----------------------------------------------------------------
% New commands etc.
\input{sl_definitions.tex}
\input{sl_symbols.tex}
%matrices
\newcommand{\I}{\mathbf{I}}
%vec of ones
%equilibrium distribution
\newcommand{\pr}{\mathbf{p}}
\newcommand{\eq}{\pr^\infty}
%other symbols
\newcommand{\w}{\mathbf{w}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\frg}{\W^\mathrm{F}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\F}{\boldsymbol{\Phi}}
\newcommand{\pot}{^{\text{pot}}}
\newcommand{\dep}{^{\text{dep}}}
\newcommand{\potdep}{^{\text{pot/dep}}}
\newcommand{\norm}{_0}
\newcommand{\inc}{_{\text{inc}}}
\newcommand{\dec}{_{\text{dec}}}
\newcommand{\incdec}{_{\text{inc/dec}}}
\newcommand{\wt}{_{\text{WT}}}
\newcommand{\ko}{_{\text{K$^\mathrm{b}$D$^\mathrm{b}$-/-}}}
\newcommand{\KO}{K$^\mathrm{b}$D$^{\mathrm{b}-/-}$}
%\newcommand{\ko}{_{\text{MHC}}}
%\newcommand{\KO}{MHC}
\newcommand{\tpre}{t_{\text{pre}}}
\newcommand{\ttrain}{t_{\text{train}}}
\newcommand{\lmax}{_{\text{max}}}
\newcommand{\lmin}{_{\text{min}}}
% ----------------------------------------------------------------
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Title info:
\title{Understanding both enhanced and impaired learning with enhanced plasticity: a saturation hypothesis (supplementary material)}
%
%% Author List:
%%
%\author{An Author$^a$, Another Author$^b$ and Yet Another Author$^c$\\
%%
%\small{\emph{$^a$ Address 1}}\\
%\small{\emph{$^b$ Address 2}}\\
%\small{\emph{$^c$ Address 3}}\\
%}
\date{}
\DeclareCaptionLabelFormat{supp}{\textbf{#1 S#2}}
\captionsetup{labelformat=supp}

@@

\setcounter{page}{5}
\setcounter{figure}{4}
\maketitle

@@

The independence and identicalness of synapses means that the state of the system can be completely described by the probability distribution of the internal states, the row vector $\pr(t)$.

The evolution of this probability is described by a continuous time Markov matrix matrix, $\frg$:
%
\begin{equation}\label{eq:evolve}
  \diff{\pr(t)}{t} = r\pr(t)\frg,
  \qquad
  \frg = f\pot\M\pot+f\dep\M\dep-\I,
\end{equation}
%
where $\I$ is the identity matrix.
Eventually, this will settle into the equilibrium distribution:
%
\begin{equation}\label{eq:eqprob}
  \eq\frg=0.
\end{equation}
%

@@

We will look at several different models,
the serial model (see \cite{Leibold2008serial} and \autoref{fig:models}\ref{fig:serial_model}) which has only two values for the synaptic weight,
the two-state model (which can be thought of as a special case of the serial model, see \autoref{fig:models}\ref{fig:binary_model}),
the multistate model (see \cite{amit1994learning,Fusi2007multistate} and \autoref{fig:models}\ref{fig:multistate_model}) which has a linearly varying synaptic weight,
and the cascade model (see \cite{Fusi2005cascade} and \autoref{fig:models}\ref{fig:cascade_model}).
We will also look at a new, pooled resource model that we will define below in \autoref{sec:pooledmodel} and a non-uniform version of the multistate model that we will describe in \autoref{sec:nonunimodel}.

@@

This model is parameterised by the range of values, $q\in[q\lmin,q\lmax]$, for potentiation and depression.
We will use the same values for potentiation and depression in the wild-type as well as potentiation in the \KO\ models.
We will use larger values for depression in the \KO\ models.
In the main text, we consider a version of this model that has resource depletion for depression only, so that potentiation transition probabilities are unaffected by the number of potentiated synapses.
This is done by setting $q\pot\lmax=q\pot\lmin$.
The values of these parameters are listed in \autoref{tab:params}.

@@

This model is parameterised by the values of $x$ chosen for potentiation and depression.
We will use a larger value for depression in the \KO\ models.
The values used are listed in \autoref{tab:params}.

@@

We will describe the effect of training by the decrease in mean synaptic weight:
%
\begin{equation}\label{eq:learning}
  L(t) = \prn{\pr(0)-\pr(t)}\w.
\end{equation}
%
If we imagine something like an LN model for the Purkinje cell, the output would depend on some linear combination of the synaptic weights (weighted by the activities of the corresponding parallel fibres).
If we're not keeping track of synaptic identity, the most natural linear combination to use would be an equal sum of them all.
The behavioural output (VOR gain) will be some unknown, non-linear function of the synaptic weights, so the best we can hope for is to reproduce qualitative features of the experiment, such as whether learning is enhanced or impaired by the mutation or pre-training.

We will assume $f\dep\wt=f\dep\ko$.
This is because the effects of the mutation are well localised to the Purkinje cells, so the activity of the parallel and climbing fibres should not change very much.
Therefore the rates of potentiation and depression should not change very much either.
This implies that the mean synaptic weight in equilibrium differs between the \KO\ models and wild type, but that is also suggested by the data regarding basal levels of phospho-GluR2 at serine 880.
As the mutation produces no change in baseline performance, there must be a compensatory mechanism somewhere else.
We will model this compensation as a simple linear offset, as could be produced by another population of neurons/synapses whose effect cancels with these neurons/synapses.

@@

The features of the actual experiments that we'd like to capture are:
\begin{enumerate}
  \item Without pre-training, gain-increase learning is significantly faster in the wild-type than in the \KO\ mice.
  \item For the wild-type, gain-increase learning is significantly faster without pre-training than with it.
  \item For the \KO\ mice, gain-increase learning is significantly faster with pre-training than without it.
  \item After pre-training, gain-increase learning is significantly faster in the \KO\ mice than in the wild-type.
%  \item Gain-decrease pre-training is slightly faster in the wild-type, but not significantly so.\label{it:gaindown}
\end{enumerate}
These questions will not be affected by any output nonlinearity, as long as it is monotonic and fixed.
We will not study gain-decrease learning, as it uses a different mechanism to gain-increase.
We are only modelling the effect of gain-decrease training on \emph{these} synapses.
We will also not discuss the curvature of the learning curves, as this can be changed by the nonlinear relation between synaptic weight and VOR gain.
%However, both the wild-type and \KO\ mice should be affected in the same way, so the difference in the curvature of the \KO\ mice and wild-type curves after pre-training would be nice to capture.

We will try to gain some analytic insight to some of these models by looking at the slope of the learning curve at the start of gain-increase training.
This is proportional to the net-flux from the states with strong synaptic weight to the weaker states, measured using the transition probabilities for gain-increase but the equilibrium distribution for either untrained or gain-decrease, assuming that pre-training lasts long enough to reach the equilibrium distribution for gain-decrease.
This should be multiplied by the difference in synaptic weights.
But, as this quantity is constant within a model and doesn't change when comparing wild-type and \KO\ models or with/without pre-training, it can be ignored.

@@

Consider the general uniform multistate model.
Then the equilibrium distribution is given by
%
\begin{equation}\label{eq:mutltieq}
  \eq_i = \frac{1-\alpha}{1-\alpha^M}\,\alpha^{i-1},
  \qquad \text{where} \quad
  \alpha=\frac{f\pot q\pot}{f\dep q\dep}.
\end{equation}
%
If we take the limit $\alpha\rightarrow1$, this becomes $\frac{1}{M}$.

The net-flux from the $\w=+1$ states to the $\w=-1$ states is:
%
\begin{equation}\label{eq:multiflux}
  \Phi = \eq_{M/2+1}f'{}\dep q\dep - \eq_{M/2}f'{}\pot q\pot = \frac{1-\alpha}{1-\alpha^M}\,\alpha^{M/2-1}\prn{\alpha-\alpha'}f'{}\dep q\dep,
\end{equation}
%
where primed values correspond to the new value of $f\pot$.

First, consider the wild-type, for which $q\pot=q\dep=q$.
Without pre-training:
%
\begin{equation}\label{eq:multiWTnopre}
  \Phi = \frac{2{\Delta f}q}{M}.
\end{equation}
%
%where it's worth remembering that $\Delta f<0$.
With pre-training:
%
\begin{equation}\label{eq:multiWTpre}
\begin{aligned}
  \Phi &= 16(\Delta f)^2q \, \frac{(1+2\Delta f)^{M/2-1} (1-2\Delta f)^{M/2-1}}
          {(1+2\Delta f)^M - (1-2\Delta f)^M} \\
       &= \frac{4{\Delta f}\,q}{M} + \CO(\Delta f)^2.
\end{aligned}
\end{equation}
%
So, we see that pre-training will speed up learning when $\Delta f$ is small (in absolute value).
%, as seen in \autoref{fig:multistate_weak}\ref{fig:multistate_weak_learnS} (black curves).
On the other hand, if $\Delta f$ is close to $\half$, pre-training will initially slow down learning a lot due to the factor of $(1-2\Delta f)^{M/2-1}$.
%, as seen in \autoref{fig:multistate_strong}\ref{fig:multistate_strong_learnS}.

@@

There are two effects here as well.
Smaller $\beta$ will increase the probability of crossing the centre of the chain, speeding up learning, but it will also concentrate probability at the ends of the chain, depleting the centre and slowing down learning.
 %(see \autoref{fig:multistate_strong}\ref{fig:multistate_strong_eq}\ref{fig:multistate_strong_eq_KO}, blue trace).
The first effect goes like $1/\beta$, whereas the second goes like $\beta^{M/2}$ and will be more significant for smaller $\beta$ or in a longer chain.
This exponential decay in the initial distribution is what allows the saturation bias to overwhelm the increases intrinsic plasticity rate and lead to impaired learning with enhanced plasticity.
It is important that a learning signal is only generated by the central transition so that the exponential decay can take effect.

@@

The results of simulations of the two-state model can be seen in
\autoref{fig:sim_results}\ref{fig:binary_sim}.
%\autoref{fig:binary}.
However, this model can be solved exactly:
%
\begin{multline}\label{eq:binarysol}
  \eq = \frac{(f\dep q\dep, f\pot q\pot)}{\lambda},
  \qquad
  \pr(t) = \eq + (\pr(t)-\eq)\e^{-\lambda rt},\\
  \qquad \text{where} \quad
  \lambda = f\pot q\pot + f\dep q\dep.
\end{multline}
%
But it is easier to just substitute $M=2$ into the formulae in \autoref{sec:multistate}.
In this case, the initial rate of change and the total change encapsulate the whole solution, as there is only a single exponential decay.

First, consider the wild-type, for which $q\pot=q\dep=q$.
Without pre-training:
%
\begin{equation}\label{eq:binWTnopre}
  \Phi = {\Delta f}\, q.
\end{equation}
%
With pre-training:
%
\begin{equation}\label{eq:binWTpre}
\begin{aligned}
  \Phi &= 2{\Delta f}\, q.
\end{aligned}
\end{equation}
%
So, we see that pre-training will always speed up learning, unlike what is seen in the experiment.
This can be seen in
\autoref{fig:sim_results}\ref{fig:binary_sim} %\autoref{fig:binary}\ref{fig:binary_learnS}
(black curves).

Now, consider the \KO\ models, for which we define $\beta=q\pot/q\dep<1$, and we set $q\pot=q$.
Without pre-training:
%
\begin{equation}\label{eq:binKOnopre}
  \Phi = \frac{2{\Delta f}\, q}{1+\beta},
\end{equation}
%
which is always larger that \eqref{eq:binWTnopre}, unlike what is seen in the experiment.
This is also seen in
\autoref{fig:sim_results}\ref{fig:binary_sim} %\autoref{fig:binary}\ref{fig:binary_learn},\ref{fig:binary_learnS}
(solid curves).
As discussed above (below \eqref{eq:multiKOnopre}), the larger value of $q\dep=q/\beta$ has two effects.
Smaller $\beta$ will increase the probability of depression, speeding up learning, but it will also decrease the probability of being ready for depression, slowing down learning.
For the serial model, we argued that the first effect goes like $1/\beta$, whereas the second goes like $\beta^{M/2}$ to leading order.
Here $M=2$, so the leading part of the two effects will cancel.
The subleading effects (the normalisation of the probabilities) are responsible for the faster learning in the \KO\ models.

@@

We made a number of assumptions in our analysis so far.
Before we declare this model to be falsified, we should think about what would happen if we relaxed these assumptions and explored a larger parameter space.

First, we assumed that the wild type had symmetric potentiation and depression.
Relaxing this assumption would correspond modelling the wild-type like the \KO\ mice, but with a different value for $\beta$.
We will still have $\beta\ko<\beta\wt$, and looking at \eqref{eq:binKOnopre} tell us that we would still have faster learning in the \KO\ models.

Secondly, we assumed that there would be equal rates of potentiation and depression in the untrained situation.
However, relaxing this assumption merely changes \eqref{eq:binKOnopre} to
%
\begin{equation}\label{eq:binKOnoprediff}
  \Phi = \frac{{\Delta f}\, q}{f\dep+\beta f\pot} = \frac{{\Delta f}\, q\pot q\dep}{f\dep q\dep+ f\pot q\pot}.
\end{equation}
%
This will not change our conclusions either.

Because for this simple binary synapse model, the ratio of the numbers of potentiated and depressed synapses is equal to the ratio of the potentiating and depressing transition rates.
This means that the leading effects of enhanced intrinsic plasticity rates and saturation bias cancel each other.
This is eve true as the number of synapses available for further plasticity approaches zero, which requires that the depressing transition be infinitely stronger than the potentiating transition, compensating for the saturation.
There is a requirement that the total number of synapses, potentiated and depressed, is fixed.
This normalization effect always dampens the effect of saturation bias, so that the effect of enhanced intrinsic plasticity rates always dominates.
Thus, we always see enhanced learning with enhanced plasticity with the binary synapse model.

As we did not need the pre-training paradigm to rule out this model, we will not look at removing the symmetry between gain-increase and decrease in detail.
However, one can see that pre-training will always increase the fraction of synapses available for depression.
This model has no mechanism that could ever result in this causing impaired learning.

@@

The equilibrium distribution, \eqref{eq:mutltieq}, still applies.
However, now the rate of change of our learning metric, \eqref{eq:learning}, will be proportional to the sum of the net fluxes between adjacent states:
%
\begin{equation}\label{eq:multiLinFlux}
  \begin{aligned}
    \Phi &= \sum_{i=1}^{M-1} \eq_{i+1} f'{}\dep q\dep - \eq_i f'{}\pot q\pot
         &= \sum_{i=1}^{M-1} \eq_i (\alpha-\alpha') f'{}\dep q\dep \\
         &= \frac{1-\alpha^{M-1}}{1-\alpha^M} \, (\alpha-\alpha') f'{}\dep q\dep.
  \end{aligned}
\end{equation}
%

First, consider the wild-type, for which $q\pot=q\dep=q$.
Without pre-training:
%
\begin{equation}\label{eq:multiLinWTnopre}
  \Phi = 2{\Delta f}\,q\,\frac{M-1}{M},
\end{equation}
%
where, once more, we should note that $\Delta f<0$.
With pre-training:
%
\begin{equation}\label{eq:multiLinWTpre}
\begin{aligned}
  \Phi &= 4{\Delta f}\, q \, \frac{(1+2\Delta f)^{M-1} - (1-2\Delta f)^{M-1}}
          {(1+2\Delta f)^M - (1-2\Delta f)^M} \\
       &= {4{\Delta f}\, q}\,\frac{M-1}{M} + \CO(\Delta f)^2.
\end{aligned}
\end{equation}
%

>>>>

% -*- TeX -*- -*- UK -*-
% ----------------------------------------------------------------
\documentclass[10pt]{article}
%\usepackage[printfigures]{figcaps}
%\usepackage[nolists,nomarkers]{endfloat}
%Preamble:
\usepackage{a4wide}
\input{sl_preamble.tex}
\hypersetup{colorlinks,linkcolor=blue,citecolor=teal,urlcolor=teal}
\input{sl_graphics_preamble.tex}
\graphicspath{{"Figures/"}}
%
\usepackage{array}
\newcounter{tablerow}[table]
\newcommand{\tablerowautorefname}{row}
\newenvironment{tabularn}[1]{\begin{tabular}{#1>{\refstepcounter{tablerow}}l}}{\end{tabular}}
%\usepackage{etoolbox}
%\AtBeginEnvironment{tabular}{\setcounter{tablerow}{0}}
\renewcommand{\figureautorefname}{Supplementary Figure}
% ----------------------------------------------------------------
% New commands etc.
\input{sl_definitions.tex}
\input{sl_symbols.tex}
%matrices
\newcommand{\I}{\mathbf{I}}
%vec of ones
%equilibrium distribution
\newcommand{\pr}{\mathbf{p}}
\newcommand{\eq}{\pr^\infty}
%other symbols
\newcommand{\w}{\mathbf{w}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\frg}{\W^\mathrm{F}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\F}{\boldsymbol{\Phi}}
\newcommand{\pot}{^{\text{pot}}}
\newcommand{\dep}{^{\text{dep}}}
\newcommand{\potdep}{^{\text{pot/dep}}}
\newcommand{\norm}{_0}
\newcommand{\inc}{_{\text{inc}}}
\newcommand{\dec}{_{\text{dec}}}
\newcommand{\incdec}{_{\text{inc/dec}}}
\newcommand{\wt}{_{\text{WT}}}
\newcommand{\ko}{_{\text{K$^\mathrm{b}$D$^\mathrm{b}$-/-}}}
\newcommand{\KO}{K$^\mathrm{b}$D$^{\mathrm{b}-/-}$}
%\newcommand{\ko}{_{\text{MHC}}}
%\newcommand{\KO}{MHC}
\newcommand{\tpre}{t_{\text{pre}}}
\newcommand{\ttrain}{t_{\text{train}}}
%\newcommand{\lmax}{_{\text{max}}}
%\newcommand{\lmin}{_{\text{min}}}
\newcommand{\modelfig}[1][a]{Figure 4#1 of the main text}
% ----------------------------------------------------------------
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Title info:
\title{Understanding both enhanced and impaired learning with enhanced plasticity: a saturation hypothesis (supplementary material)}
%
%% Author List:
%%
%\author{An Author$^a$, Another Author$^b$ and Yet Another Author$^c$\\
%%
%\small{\emph{$^a$ Address 1}}\\
%\small{\emph{$^b$ Address 2}}\\
%\small{\emph{$^c$ Address 3}}\\
%}
\date{}
\DeclareCaptionLabelFormat{supp}{\textbf{#1 S#2}}
\captionsetup{labelformat=supp}

@@

\setcounter{page}{8}
\setcounter{figure}{7}
\maketitle

@@

The independence and identicalness of synapses means that the state of the system can be completely described by the probability distribution over the internal states, the \emph{row} vector $\pr(t)$ which evolves as
%
\begin{equation}\label{eq:evolve}
  \diff{\pr(t)}{t} = r\pr(t)\frg,
  \qquad
  \frg = f\pot\M\pot+f\dep\M\dep-\I,
\end{equation}
%
where $\frg$ is a continuous time Markov matrix matrix and $\I$ is the identity matrix.
Eventually, this will settle into the equilibrium distribution:
%
\begin{equation}\label{eq:eqprob}
  \eq\frg = \mathbf{0}.
\end{equation}
%

@@

We will look at several different models,
the serial model (see \cite{Leibold2008serial,Ben-DayanRubin2007sparse} and \autoref{fig:models}\ref{fig:serial_model}) which has only two values for the synaptic weight,
the two-state model (which can be thought of as a special case of the serial model, see \autoref{fig:models}\ref{fig:binary_model}),
the multistate model (see \cite{amit1994learning,Fusi2007multistate} and \autoref{fig:models}\ref{fig:multistate_model}) which has a linearly varying synaptic weight,
and the cascade model (see \cite{Fusi2005cascade} and \autoref{fig:models}\ref{fig:cascade_model}).
We will also look at a new, pooled resource model that we will define below in \autoref{sec:pooledmodel} and a non-uniform version of the multistate model that we will describe in \autoref{sec:nonunimodel}.

@@

This model is parameterised by the range of values, $q\in[q\lmin,q\lmax]$, for potentiation and depression.
We will use the same values for potentiation and depression in the wild-type as well as potentiation in the \KO\ models.
We will use larger values for depression in the \KO\ models.
In the main text, we consider a version of this model that has resource depletion for depression only, so that potentiation transition probabilities are unaffected by the number of potentiated synapses.
This is done by setting $q\pot\lmax=q\pot\lmin$.
The values of these parameters are listed in \autoref{fig:sim_results}\ref{tab:params}.

@@

This model is parameterised by the values of $x$ chosen for potentiation and depression.
We will use a larger value for depression in the \KO\ models.
The values used are listed in \autoref{fig:sim_results}\ref{tab:params}.

@@

We will describe the effect of training by the decrease in mean synaptic weight:
%
\begin{equation}\label{eq:learning}
  L(t) = \prn{\pr(0)-\pr(t)}\w.
\end{equation}
%
One can approximate the input to a Purkinje cell as some linear combination of the synaptic weights (weighted by the activities of the corresponding parallel fibres).
If we are not keeping track of synaptic identity, the most natural linear combination to use would be an equal sum of them all.
The behavioural output (VOR gain) will be some unknown, non-linear function of the synaptic weights, so the best we can hope for is to reproduce qualitative features of the experiment, such as whether learning is enhanced or impaired by the mutation or pre-training.

As the knockout produces no change in baseline performance, there must be a compensatory mechanism somewhere else.
We will model this compensation as a simple linear offset, as could be produced by another population of neurons/synapses whose effect cancels with these neurons/synapses.

We will assume $f\dep\wt=f\dep\ko$.
This is because the relevant effects of the knockout are well localised to the Purkinje cells, as shown by the rescue data, so the activity of the parallel and climbing fibres should not change very much.
Therefore the rates of potentiation and depression should not change very much either.
%This implies that the mean synaptic weight in equilibrium differs between the \KO\ models and wild type, but that is also suggested by the data regarding basal levels of phospho-GluR2 at serine 880.

@@

The features of the experiments shown in \modelfig\ that we'd like to capture are:
%
\begin{enumerate}
  \item Without pre-training, gain-increase learning is significantly faster in the wild-type than in the \KO\ mice.
  \item For the wild-type, gain-increase learning is significantly faster without pre-training than with it.
  \item For the \KO\ mice, gain-increase learning is significantly faster with pre-training than without it.
  \item After pre-training, gain-increase learning is significantly faster in the \KO\ mice than in the wild-type.
%  \item Gain-decrease pre-training is slightly faster in the wild-type, but not significantly so.\label{it:gaindown}
\end{enumerate}
%
These questions will not be affected by any output nonlinearity, as long as it is monotonic and fixed.
We will not study gain-decrease learning, as it uses a different mechanism to gain-increase.
We are only modelling the effect of gain-decrease training on \emph{these} synapses.
We will also not discuss the curvature of the learning curves, as this can be changed by the nonlinear relation between synaptic weight and VOR gain.
%However, both the wild-type and \KO\ mice should be affected in the same way, so the difference in the curvature of the \KO\ mice and wild-type curves after pre-training would be nice to capture.

We will try to gain some analytic insight to some of these models by looking at the slope of the learning curve at the start of gain-increase training.
This is proportional to the net-flux from the states with strong synaptic weight to the weaker states, measured using the transition rates for gain-increase but the equilibrium distribution for either untrained or gain-decrease, assuming that pre-training lasts long enough to reach the equilibrium distribution for gain-decrease.
%This should be multiplied by the difference in synaptic weights.
%But, as this quantity is constant within a model and doesn't change when comparing wild-type and \KO\ models or with/without pre-training, it can be ignored.

@@

Consider the general uniform serial model.
Then the equilibrium distribution is given by
%
\begin{equation}\label{eq:mutltieq}
  \eq_i = \frac{1-\alpha}{1-\alpha^M}\,\alpha^{i-1},
  \qquad \text{where} \quad
  \alpha=\frac{f\pot q\pot}{f\dep q\dep}.
\end{equation}
%
If we take the limit $\alpha\rightarrow1$, this becomes $\frac{1}{M}$.

The net-flux from the $\w=+1$ states to the $\w=-1$ states is:
%
\begin{equation}\label{eq:multiflux}
  \Phi = \eq_{M/2+1}f'{}\dep q\dep - \eq_{M/2}f'{}\pot q\pot = \frac{1-\alpha}{1-\alpha^M} \, \alpha^{M/2-1} \prn{\frac{\alpha}{\alpha\inc} - 1} f\inc\pot  q\pot,
\end{equation}
%
where $\alpha\inc$ is $\alpha$, but with $f\potdep$ replaced by $f\potdep\inc$.

To see the effects of the knockout, we can see if this increases or decreases with $q\dep$:
%
\begin{equation}\label{eq:serialderq}
  q\dep \pdiff{\Phi}{q\dep} = 
    \frac{\alpha^{\frac{M}{2} - 1}}{(1 - \alpha^M)^2} 
    \brk{(1 - \alpha^M) - \frac{M}{2} (1 - \alpha) (1 + \alpha^M) } 
    \prn{\frac{\alpha}{\alpha\inc} - 1} f\inc\pot  q\pot.
\end{equation}
%
This quantity can have either sign, depending on $\alpha$ and $M$, so $\Phi$ is not a monotonic function of $q\dep$.
Therefore, enhancing plasticity in this way can either enhance or impair the initial learning rate, depending on the parameters.

The effect of pre-training is to replace $f\potdep$ with $f\potdep\dec$ without changing $f\potdep\inc$, which increases $f\pot$ and decreases $f\dep$.
Therefore, we can see the effects of pre-training on the initial learning rate by seeing if it increases or decreases with $f\pot$ (with $f\dep = 1 - f\pot$):
%
\begin{equation}\label{eq:serialderf}
  f\pot f\dep \pdiff{\Phi}{f\pot} =  
    \frac{\alpha^{\frac{M}{2} - 1}}{(1 - \alpha^M)^2} 
      \brk{ \prn{\frac{\alpha}{\alpha\inc} - 1} \frac{M}{2} (1 - \alpha) (1 + \alpha^M) 
      - \prn{\frac{\alpha^2}{\alpha\inc} - 1} (1 - \alpha^M) }  
      f\inc\pot  q\pot.
\end{equation}
%
This quantity can have either sign, depending on $\alpha$ and $M$, so $\Phi$ is not a monotonic function of $f\pot$.
Therefore, pre-training can either enhance or impair the initial learning rate, depending on the parameters.

The lack of monotonicity shown above allows this model to reproduce the key qualitative features of the experiments outlined in \autoref{sec:results}.
This can be illustrated by looking at a special case, where $q\pot = q\dep = q$ for the wild type, $f\dep_0 = \frac{1}{2}$, $f\dep\dec = \frac{1}{2} - \Delta f$ and $f\dep\inc = \frac{1}{2} - \Delta f$.

First, consider the wild-type, for which $q\pot=q\dep=q$.
Without pre-training:
%
\begin{equation}\label{eq:multiWTnopre}
  \Phi = \frac{2{\Delta f}q}{M}.
\end{equation}
%
%where it's worth remembering that $\Delta f<0$.
With pre-training:
%
\begin{equation}\label{eq:multiWTpre}
\begin{aligned}
  \Phi &= 16(\Delta f)^2q \, \frac{(1+2\Delta f)^{M/2-1} (1-2\Delta f)^{M/2-1}}
          {(1+2\Delta f)^M - (1-2\Delta f)^M} \\
       &= \frac{4{\Delta f}\,q}{M} + \CO(\Delta f)^2.
\end{aligned}
\end{equation}
%
So, we see that pre-training will speed up learning when $\Delta f$ is small.
%, as seen in \autoref{fig:multistate_weak}\ref{fig:multistate_weak_learnS} (black curves).
On the other hand, if $\Delta f$ is close to $\half$, pre-training will initially slow down learning a lot due to the factor of $(1-2\Delta f)^{M/2-1}$.
%, as seen in \autoref{fig:multistate_strong}\ref{fig:multistate_strong_learnS}.

@@

There are two effects here as well.
Smaller $\beta$ will increase the probability of crossing the centre of the chain, speeding up learning, but it will also concentrate probability at the ends of the chain, depleting the centre and slowing down learning.
 %(see \autoref{fig:multistate_strong}\ref{fig:multistate_strong_eq}\ref{fig:multistate_strong_eq_KO}, blue trace).
The first effect goes like $1/\beta$, whereas the second goes like $\beta^{M/2}$ and will be more significant for smaller $\beta$ or in a longer chain.
This exponential decay in the initial distribution is what allows the saturation bias to overwhelm the increased intrinsic plasticity rate and lead to impaired learning with enhanced plasticity.
It is important that a learning signal is only generated by the central transition so that the exponential decay can take effect.

@@

The results of simulations of the two-state model can be seen in
\autoref{fig:sim_results}\ref{fig:binary_sim}.
%\autoref{fig:binary}.
However, this model can be solved exactly:
%
\begin{multline}\label{eq:binarysol}
  \eq = \frac{(f\dep q\dep, f\pot q\pot)}{\lambda},
  \qquad
  \pr(t) = \eq + (\pr(0)-\eq)\e^{-\lambda rt},\\
  \qquad \text{where} \quad
  \lambda = f\pot q\pot + f\dep q\dep.
\end{multline}
%
But it is easier to just substitute $M=2$ into the formulae in \autoref{sec:multistate}.
%In this case, the initial rate of change and the total change encapsulate the whole solution, as there is only a single exponential decay.

We find that
%
\begin{equation}\label{eq:binaryflux}
\begin{aligned}
  \Phi &= \frac{q\pot q\dep \Delta f}{f\pot q\pot + f\dep q\dep}, &\qquad
  \pdiff{\Phi}{q\dep} &= \frac{f\pot (q\pot)^2 \Delta f}{(f\pot q\pot + f\dep q\dep)^2}, \\ &&
  \pdiff{\Phi}{f\pot} &= \frac{(f\pot\inc q\pot + f\dep\inc q\dep) q\pot q\dep}{(f\pot q\pot + f\dep q\dep)^2}.
\end{aligned}
\end{equation}
%
The flux is a monotonically increasing function of $q\dep$ and $f\pot$.
Therefore, enhancing plasticity in this way or pre-training can only enhance the initial learning rate, unlike the experimental results shown in \modelfig.

The monotonicity shown above means that this model cannot reproduce the key qualitative features of the experiments outlined in \autoref{sec:results}, at least as for the initial learning rate.
This can be illustrated by looking at a special case, where $q\pot = q\dep = q$ for the wild type, $f\dep_0 = \frac{1}{2}$, $f\dep\dec = \frac{1}{2} - \Delta f$ and $f\dep\inc = \frac{1}{2} - \Delta f$.

First, consider the wild-type, for which $q\pot=q\dep=q$.
Without pre-training:
%
\begin{equation}\label{eq:binWTnopre}
  \Phi = {\Delta f}\, q.
\end{equation}
%
With pre-training:
%
\begin{equation}\label{eq:binWTpre}
\begin{aligned}
  \Phi &= 2{\Delta f}\, q.
\end{aligned}
\end{equation}
%
So, we see that pre-training will speed up learning for all parameter values, unlike what is seen in the experiment.
This can be seen in
\autoref{fig:sim_results}\ref{fig:binary_sim} %\autoref{fig:binary}\ref{fig:binary_learnS}
(black curves).

Now, consider the \KO\ models, for which we define $\beta=q\pot/q\dep<1$, and we set $q\pot=q$.
Without pre-training:
%
\begin{equation}\label{eq:binKOnopre}
  \Phi = \frac{2{\Delta f}\, q}{1+\beta},
\end{equation}
%
which is always larger than \eqref{eq:binWTnopre}, unlike what is seen in the experiments (see \modelfig).
This is also seen in
\autoref{fig:sim_results}\ref{fig:binary_sim} %\autoref{fig:binary}\ref{fig:binary_learn},\ref{fig:binary_learnS}
(solid curves).
As discussed above (below \eqref{eq:multiKOnopre}), the larger value of $q\dep=q/\beta$ has two effects.
Smaller $\beta$ will increase the probability of depression, speeding up learning, but it will also decrease the probability of being ready for depression, slowing down learning.
For the serial model, we argued that the first effect goes like $1/\beta$, whereas the second goes like $\beta^{M/2}$ to leading order.
Here $M=2$, so the leading part of the two effects will cancel.
The subleading effects (the normalisation of the probabilities) are responsible for the faster learning in the \KO\ models.

@@

For this simple binary synapse model, the ratio of the numbers of potentiated and depressed synapses is equal to the ratio of the potentiating and depressing transition rates.
This means that the leading effects of enhanced intrinsic plasticity rates and saturation bias cancel each other.
This is even true as the number of synapses available for further plasticity approaches zero, which requires that the depressing transition be infinitely stronger than the potentiating transition, compensating for the saturation.
There is a requirement that the total number of synapses, potentiated and depressed, is fixed.
This normalization effect always dampens the effect of saturation bias, so that the effect of enhanced intrinsic plasticity rates always dominates.
Thus, we always see enhanced learning with enhanced plasticity with the binary synapse model.

One can see that pre-training will always increase the fraction of synapses available for depression.
This model has no mechanism that could ever result in this causing impaired learning.

@@

The equilibrium distribution, \eqref{eq:mutltieq}, still applies.
However, now the rate of change of our learning metric, \eqref{eq:learning}, will be proportional to the sum of the net fluxes between adjacent states:
%
\begin{equation}\label{eq:multiLinFlux}
  \begin{aligned}
    \Phi &= \sum_{i=1}^{M-1} \eq_{i+1} f'{}\dep q\dep - \eq_i f'{}\pot q\pot
         %&= \sum_{i=1}^{M-1} \eq_i (\alpha-\alpha') f'{}\dep q\dep \\
         &= \frac{1-\alpha^{M-1}}{1-\alpha^M} \, \prn{\frac{\alpha}{\alpha\inc}} f\inc\pot q\pot.
  \end{aligned}
\end{equation}
%

To see the effects of the knockout, we can see if this increases or decreases with $q\dep$:
%
\begin{equation}\label{eq:multiderq}
  q\dep \pdiff{\Phi}{q\dep} = 
    \frac{\alpha^{M - 1}}{(1 - \alpha^M)^2}
    \brk{1 - \alpha}
    \brk{M - \sum_{i=1}^M \alpha^{i-1}} 
    \prn{\frac{\alpha}{\alpha\inc} - 1} f\inc\pot  q\pot.
\end{equation}
%
If $\alpha < 1$, both quantities in square brackets are positive, as the sum contains $M$ terms that are all at most $1$.
If $\alpha > 1$, both of them are negative.
Therefore this quantity is positive, so $\Phi$ is a monotonically increasing function of $q\dep$.
Therefore, enhancing plasticity in this way can only enhance the initial learning rate, for all values the parameters.
